Exponential Random Graph Models with Big Networks: Maximum Pseudolikelihood Estimation and the Parametric Bootstrap

### 2017-10-05 TeX/complenet18.tex
Reduced CompleNet paper to 15 pages

\section{Introduction}
The field of network science faces a double-edge sword when it comes to computationally intensive research. First, the availability of digital source data has led the growth in network science to be synonymous with the growth in research on big data.  Faraj et al. (2008, pp. 19--20) \cite{faraj2008electronic} notes that, \begin{quote} ``In recent years the proliferation of advanced Information Technology has not only facilitated the collection of large-scale network data but also increased the availability of large-scale network data analysis tools and techniques. This has led to an explosion of interest in large network research in fields spanning biology, physics, mathematics, and social sciences.''  \end{quote} Second, analytical methods are growing more sophisticated, increasingly involving iterative and/or simulation-based optimization, rather than simple descriptive calculations \cite{snijders2011statistical}.  Closing the gap in terms of the size of the networks to which it is feasible to apply the most sophisticated methods of network modeling requires research into scalable methods of inference. We propose a method of statistical inference for one of the most popular models for networks---the exponential random graph model (ERGM), in which both parameter estimates and confidence intervals are derived, that can require less than half the compute time of currently used methods.

@@

as the set of all possible networks on $N$ nodes without loops. Note that the cardinality of set $\mathcal{G}(N)$ is increasing exponentially for every newly included actor, which results in $2^{N(N-1)/2}$ total elements. For even a small number of nodes, the cardinality of $\mathcal{G}(N)$ turns out to be an astronomically large number. For this reason calculating the likelihood function of the ERGM, which requires evaluating a normalizing constant on $\mathcal{G}(N)$  is either extremely time-consuming or with today's technology not achievable. As a consequence, many approximation methods have been provided by the literature, with the most popular method making use of Markov Chain Monte Carlo (MCMC) methods \cite{hunter2012computational}, as we will introduce in the next section.\\  

@@

where $\theta \in \mathbb{R}^q$ is a $q-$dimensional vector of parameters, $\Gamma:\mathcal{G}(N) \to \mathbb{R}^q~,~G \mapsto (\Gamma_1(G),\dots,\Gamma_q(G))^T$ is a $q$-dimensional function of different network statistics and $c(\theta):= \sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))$ is a normalization constant which ensures that (\ref{ERGM}) defines a probability function on $\mathcal{G}$.

@@

As already mentioned, a specific network $G$ can be considered as a manifestation of a matrix-like random variable, whose probability of occurrence can be modeled with equation (\ref{ERGM}). The generative processes captured by a model (e.g., density, reciprocity, popularity, clustering) are informed by the decision regarding which network statistics (i.e., $\Gamma(\cdot)$) are incorporated (see Snijders et al. \cite{SnijdersTomA.B..2006} for a detailed discussion). The flexibility of the ERGM in capturing virtually any network generative process has led to it being applied broadly across several fields, including (but not limited to) sociology \cite{smith2016ethnic}%,srivastava2011culture}

@@

\section{Estimation}
As mentioned above calculating $c(\theta)$ is not achievable for most cases with today's technology.  Straightforward application of either maximum likelihood estimation and Bayesian inference are, therefore, not possible. The first method proposed in the literature for estimating ERGM parameters was maximum pseudolikelihood estimation \cite{StraussIkeda1990}. Under maximum pseudolikelihood estimation (MPLE), the joint distribution is replaced by the product over conditional distributions \cite{besag1986statistical}.  
\begin{equation*}
\mathbb{P}_{\theta}(G) \approx \prod_{ij} \mathbb{P}_{\theta}\left(g_{ij}|G_{-ij}\right), 
%\label{ILogit}
\end{equation*}
where $G_{-ij}$ is the adjacency matrix, excluding element $ij$. The conditional probability of a tie in ERGM reduces, conveniently, to a logistic regression form given by 
\begin{equation*}
\mathbb{P}_{\theta}\left(g_{ij}=1| G_{-ij} \right) = \text{\\} \text{logit}^{-1}\left( \theta^T \cdot \delta(\Gamma(G))  \right), 
%\label{ILogit}
\end{equation*}
where $\delta(\Gamma(G))$ is the ``change statistic'' given by the difference in the network statistics when the $ij$ element is toggled from 0 to 1 (i.e., $\Gamma(G|g_{ij}=1))-\Gamma(G|g_{ij}=0)$), and $\text{logit}^{-1}(x)=1/(1+\exp(-x))$ \cite{goodreau2009birds}. For the ERGM, the pseudolikelihood function can be maximized using logistic regression software, in which the dependent variable is given by the elements of the adjacency matrix, and the covariates are given by the values of the change statistics corresponding to each element of the adjacency matrix.

@@

\noindent MCMC methods are used to simulate the $L$ networks. As we demonstrate below, the MCMLE grows more accurate as $L$ increases. Indeed, MCMLE approaches the MLE as the number of networks simulated goes to infinity. Snijders (2002) \cite{snijders2002markov} provides a Metropolis Hastings algorithm to simulate networks:\\[0.2cm]

@@

%Choose a matrix $G^{(0)} \in \mathcal{G}(N)$ to start with (e.g., the observed network). For $k \in \{0,...,L-1\}$ recursively proceed as follows:\\
%\begin{enumerate}
%\item Randomly choose an edge $(i,j)$ where $i \neq j$ from $G^{(k)}$
%\item Compute the value
%$$\pi := \dfrac{\mathbb{P}_{\theta}(Y_{ij} \neq g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}{\mathbb{P}_{\theta}(Y_{ij}= g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}$$
%\item Fix $\delta:= \min\{1, \pi\}$ and draw a random number $Z$ from Bin$(1, \delta)$. If
%\begin{itemize}
%\item $Z=0$, let $G^{(k+1)} := G^{(k)}$ 
%\item $Z=1$, define $G^{(k+1)}$ via
%$$g_{pq}^{(k+1)}=\begin{cases}
%1-g_{pq}^{(k)}& \text{if}~ (p,q)=(i,j) \\
%g_{pq}^{(k)} &\text{if}~ (p,q) \neq (i,j) 
%\end{cases}$$
%\end{itemize}
%\item Start at step 1 with $G^{(k+1)}$.
%\end{enumerate}
\vspace{0.3cm}
\noindent The depicted algorithm provides a sequence of random networks $G_(0),\dots,G_(L)$ via a Metropolis-Hastings sampler \cite{chib1995understanding}. Since the original matrix was chosen randomly and the first simulated networks are very dependent on the chosen matrix (only one edge is changed per iteration), usually
the first $B$ networks, where $N << B << L$, are discarded as the so called Burn-In.

>>>>

\section{Introduction}
The field of network science faces a double-edge sword when it comes to computationally intensive research. First, the availability of digital source data has led the growth in network science to be synonymous with the growth in research on big data. % Faraj et al. (2008, pp. 19--20) \cite{faraj2008electronic} notes that, \begin{quote} ``In recent years the proliferation of advanced Information Technology has not only facilitated the collection of large-scale network data but also increased the availability of large-scale network data analysis tools and techniques. This has led to an explosion of interest in large network research in fields spanning biology, physics, mathematics, and social sciences.''  \end{quote} 
Second, analytical methods are growing more sophisticated, increasingly involving iterative and/or simulation-based optimization, rather than simple descriptive calculations \cite{snijders2011statistical}.  Closing the gap in terms of the size of the networks to which it is feasible to apply the most sophisticated methods of network modeling requires research into scalable methods of inference. We propose a method of statistical inference for one of the most popular models for networks---the exponential random graph model (ERGM), in which both parameter estimates and confidence intervals are derived, that can require less than half the compute time of currently used methods.

@@

as the set of all possible networks on $N$ nodes without loops. Note that the cardinality of set $\mathcal{G}(N)$ is increasing exponentially for every newly included actor, which results in $2^{N(N-1)/2}$ total elements. %For even a small number of nodes, the cardinality of $\mathcal{G}(N)$ turns out to be an astronomically large number. 
For this reason calculating the likelihood function of the ERGM, which requires evaluating a normalizing constant on $\mathcal{G}(N)$  is either extremely time-consuming or with today's technology not achievable. As a consequence, many approximation methods have been provided by the literature, with the most popular method making use of Markov Chain Monte Carlo (MCMC) methods \cite{hunter2012computational}, as we will introduce in the next section.\\  

@@

where $\theta \in \mathbb{R}^q$ is a $q-$dimensional vector of parameters, $\Gamma:\mathcal{G}(N) \to \mathbb{R}^q~,~G \mapsto (\Gamma_1(G),\dots,\Gamma_q(G))^T$ is a $q$-dimensional function of different network statistics and $c(\theta):= \sum_{G^* \in \mathcal{G}(N)} \exp(\theta^T \cdot \Gamma(G^*))$ is a normalization constant which ensures that (\ref{ERGM}) defines a probability function on $\mathcal{G}(N)$.

@@

%As already mentioned, a specific network $G$ can be considered as a manifestation of a matrix-like random variable, whose probability of occurrence can be modeled with equation (\ref{ERGM}). 
The generative processes captured by a model (e.g., density, reciprocity, popularity, clustering) are informed by the decision regarding which network statistics (i.e., $\Gamma(\cdot)$) are incorporated (see Snijders et al. \cite{SnijdersTomA.B..2006} for a detailed discussion). The flexibility of the ERGM in capturing virtually any network generative process has led to it being applied broadly across several fields, including (but not limited to) sociology \cite{smith2016ethnic}%,srivastava2011culture}

@@

\section{Estimation}
As mentioned above calculating $c(\theta)$ is not achievable for most cases with today's technology.  Straightforward application of either maximum likelihood estimation and Bayesian inference are, therefore, not possible. The first method proposed in the literature for estimating ERGM parameters was maximum pseudolikelihood estimation \cite{StraussIkeda1990}. Under maximum pseudolikelihood estimation (MPLE), the joint distribution is replaced by the product over conditional distributions \cite{besag1986statistical}.  
\mathbb{P}_{\theta}(G) \approx \prod_{ij} \mathbb{P}_{\theta}\left(g_{ij}|G_{-ij}\right), 
%\label{ILogit}
where $G_{-ij}$ is the adjacency matrix, excluding element $ij$. The conditional probability of a tie in ERGM reduces, conveniently, to a logistic regression form given by 
\mathbb{P}_{\theta}\left(g_{ij}=1| G_{-ij} \right) = \text{\\} \text{logit}^{-1}\left( \theta^T \cdot \delta(\Gamma(G))  \right), 
%\label{ILogit}
where $\delta(\Gamma(G))$ is the ``change statistic'' given by the difference in the network statistics when the $ij$ element is toggled from 0 to 1 (i.e., $\Gamma(G|g_{ij}=1))-\Gamma(G|g_{ij}=0)$), and $\text{logit}^{-1}(x)=1/(1+\exp(-x))$ \cite{goodreau2009birds}. For the ERGM, the pseudolikelihood function can be maximized using logistic regression software, in which the dependent variable is given by the elements of the adjacency matrix, and the covariates are given by the values of the change statistics corresponding to each element of the adjacency matrix.

@@

\noindent MCMC methods are used to simulate the $L$ networks. As we demonstrate below, the MCMLE grows more accurate as $L$ increases. Indeed, MCMLE approaches the MLE as the number of networks simulated goes to infinity. Snijders (2002) \cite{snijders2002markov} provides a Metropolis Hastings algorithm \cite{chib1995understanding} to simulate networks:\\[0.2cm]

@@

%Choose a matrix $G^{(0)} \in \mathcal{G}(N)$ to start with (e.g., the observed network). For $k \in \{0,...,L-1\}$ recursively proceed as follows:\\
%\begin{enumerate}
%\item Randomly choose an edge $(i,j)$ where $i \neq j$ from $G^{(k)}$
%\item Compute the value
%$$\pi := \dfrac{\mathbb{P}_{\theta}(Y_{ij} \neq g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}{\mathbb{P}_{\theta}(Y_{ij}= g_{ij}^{(k)} | Y_{ij}^c=G_{ij}^c)}$$
%\item Fix $\delta:= \min\{1, \pi\}$ and draw a random number $Z$ from Bin$(1, \delta)$. If
%\begin{itemize}
%\item $Z=0$, let $G^{(k+1)} := G^{(k)}$ 
%\item $Z=1$, define $G^{(k+1)}$ via
%$$g_{pq}^{(k+1)}=\begin{cases}
%1-g_{pq}^{(k)}& \text{if}~ (p,q)=(i,j) \\
%g_{pq}^{(k)} &\text{if}~ (p,q) \neq (i,j) 
%\end{cases}$$
%\end{itemize}
%\item Start at step 1 with $G^{(k+1)}$.
%\end{enumerate}
%\vspace{0.3cm}
%\noindent The depicted algorithm provides a sequence of random networks $G_(0),\dots,G_(L)$ via a Metropolis-Hastings sampler \cite{chib1995understanding}. Since the original matrix was chosen randomly and the first simulated networks are very dependent on the chosen matrix (only one edge is changed per iteration), usually
%the first $B$ networks, where $N << B << L$, are discarded as the so called Burn-In.
